openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2022-03-28 paper/main.tex
Update benchmarks section

% General problem statement
One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data points~\citep{becht2019dimensionality}.
The culprit for slow response time stems from
particular optimization procedures and their specific implementation in popular
open-source libraries. Until quite recently, most popular implementations of
t-SNE were based on either a direct implementation of the t-SNE algorithm with
asymptotic time complexity $\mathcal{O}(N^2)$, or the Barnes-Hut approximation
scheme developed by
\citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}
 (N \log N)$, where $N$ is the number of data items ({\em e.g.} cells).

% Problem with Python
In Python, the most widely-used implementation of t-SNE came from \pkg{scikit-learn},
which exhibits long runtimes when compared to its C++ counterpart --
\pkg{MulticoreTSNE}. The multi-threaded \pkg{MulticoreTSNE}
implementation~\citep{Ulyanov2016} can construct t-SNE embeddings of millions of
data points in a matter of hours (Fig.~\ref{fig:benchmarks_py}).
However, recently, \citet{linderman2019fast} developed a
new approximation scheme -- FIt-SNE -- which futher reduces the asymptotic time
complexity to $\mathcal{O}(N)$. This approximation scheme is included in \pkg{FIt-SNE} and \opentsne, enabling the embedding of large data sets in a matter
of minutes.

% We benchmark Python
Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python t-SNE implementations,
including those from \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1),
\pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.1). Benchmarks were performed on
an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original
t-SNE parameters, as some implementations do not allow for their modification.

To ensure a realistic benchmarking scenario, we utilize the 10X
Genomics 1.3 million mouse brain data set, containing 1.3 million single-cell 
gene-expression profiles. We preprocess the data using the standard single-cell
analysis pipeline and extract the top 50 principal components. To generate
benchmark data sets of different sizes, we subsample the data set five times,
resulting in five different data sets for each sample size. In total, we run
each implementation on 30 different data sets.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{benchmarks_python-final}
  \caption{\label{fig:benchmarks_py}
  We benchmark \opentsne\ (v0.6.1) against three popular open-source
  implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
  (v1.0.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
  on a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
  that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
  same interpolation-based approximation scheme, while \pkg{scikit-learn} and
  \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
}
\end{figure*}

% Problem with other programming languages
In R, the only widely used implementation of t-SNE -- \pkg{Rtsne}~\citep{krijthe2015rtsne} -- is based on the Barnes-Hut approximation, and is closely comparable to \pkg{MulticoreTSNE}. In Julia, the only available implementation -- \pkg{TSne.jl} -- is based on a direct $\mathcal{O}(N^2)$ implementation of the algorithm, which allows it to be used only on smaller data sets containing thousands of data points.
Matlab is proprietary software, but also implements a naive version of the t-SNE algorithm, therefore we would expect its performance to mirror that of \pkg{TSne.jl} from Julia.

% We benchmark other programming languages
Using the same benchmarking setup as before, we ...
The results are shown in Fig.~\ref{fig:benchmarks_lang}. 
Using 8 cores, Rtsne runs out of stack memory at 500k

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{benchmarks_langs-final}
  %\includegraphics[width=\textwidth]{benchmarks-combined}
  \caption{\label{fig:benchmarks_lang}
  We choose the fastest available, open-source implementations of the t-SNE algorithm from the Python, R, and Julia programming languages. We choose \opentsne\ for Python, \pkg{Rtsne} for R, and \pkg{TSne.jl} for Julia. The benchmarks reflect the time complexity of the implemented algorithms. For instance, \pkg{TSne.jl} implements a naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all but the smallest data sets. \pkg{Rtsne} scales similarily to \pkg{MulticoreTSNE} in Python, as it implements the $\mathcal{O}(N \log N)$ Barnes-Hut approximation scheme. Interestingly, the multi-threaded implementation version of \pkg{Rtsne} crashes on benchmark data sets containing over 500,000 data points due to lack of memory. This would seriously hamper its utility on consumer-grade platforms, as our benchmark platform contained 128 GB of memory. \opentsne\ is the only library implementing the more recent FIt-SNE approximation scheme, which makes it suitable for even the largest data sets.
}
\end{figure*}

Benchmark results (Fig.~\ref{fig:benchmarks_py}) confirm that both \pkg{FIt-SNE}
and \opentsne\ scale better than their Barnes-Hut counterparts --
\pkg{scikit-learn} and \pkg{MulticoreTSNE}. While \opentsne's implementation
uses Python which incurs some runtime overhead compared to its C++ counterpart
\pkg{FIt-SNE}, the two libraries' speeds are surprisingly comparable. Modern
computer processors contain multiple cores, allowing us to use multi-threading,
which further reduces the gap between \opentsne\ and \pkg{FIt-SNE}. On the
server-grade processor, \opentsne\ is slightly faster than its pure C++
counterpart when utilizing multiple cores. \opentsne\ uses \pkg{numpy} for most
linear algebra operations, which may make better use of the Intel Math Kernel
Library (MKL), which is aggressively optimized on Intel(R) Xeon(R) processors.

\opentsne\ provides a flexible API, allowing splitting up the
embedding-construction process into several parts, caching slow operations.
Running t-SNE optimization in stages enables users to quickly experiment with
different parameter settings and iterate on their final visualizations.

@@

\begin{tabular}{l c c c c c|c c}
\toprule
\setlength\tabcolsep{6pt}
& \rot{\pkg{scikit-learn}} & \rot{\pkg{BHTSNE}} & \rot{\pkg{MulticoreTSNE}} & \rot{\pkg{FIt-SNE}} & \rot{\pkg{openTSNE}} & \rot{\pkg{Rtsne}} & \rot{\pkg{Tsne.jl}} \\

\toprule
\textbf{Ease of Installation} \\
Native Python installation & \checkmark & & & & \checkmark & & \\
\pkg{PyPI} package & \checkmark & & \checkmark & & \checkmark & & \\
\pkg{conda} package & \checkmark & & & & \checkmark & & \\

\hline
\textbf{Approximation Schemes} \\
None ($\mathcal{O}(N^2)$) & & & & & & & \checkmark \\
Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \\
FIt-SNE ($\mathcal{O}(N)$) & & & & \checkmark & \checkmark & & \\

\hline
\textbf{Advanced Features and Extensions} \\
%Multiscale Gaussian kernels & & & & \checkmark & \checkmark & & \\
Extensible affinity models & & & & & \checkmark & & \\
Variable degrees of freedom & & & & \checkmark & \checkmark & & \\
Variable exaggeration & & & & \checkmark & \checkmark & & \\
Globally-consistent initialization & & & & \checkmark & \checkmark & & \\
Automatic learning rate & & & & \checkmark & \checkmark & & \\
Adding samples to embeddings & & & & & \checkmark & & \\
Callback system & & & & & \checkmark & & \\
Interactive optimization  & & & & & \checkmark & & \\

\hline
\textbf{Project Quality}\\
Actively maintained & \checkmark & & & & \checkmark & \checkmark & \checkmark \\
Continuous integration & \checkmark & & \checkmark & & \checkmark & \checkmark & \checkmark \\
User documentation & \checkmark & & & & \checkmark & \checkmark  & \\
End-to-end usage examples & \checkmark & & & \checkmark & \checkmark & \checkmark & \checkmark \\
API reference & \checkmark & & & & \checkmark & \checkmark &  \\
\bottomrule
\end{tabular}
\end{center}

\caption{\label{tab:features}
  We compare the features of \opentsne~(v0.6.1) to four popular open-source Python
  t-SNE implementations from \pkg{scikit-learn}~(v1.0.1),
  \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
  \pkg{FIt-SNE}~(v1.1.0), as well as the major implementations in R and Julia,
  \pkg{Rtsne}~(v0.15), \pkg{Tsne.jl}~(v1.3.0). Active maintenance indicates whether the project has had any meaningful development in the past year from the time of writing.
}
\end{table}

@@

\section*{Computational details}

The results in this paper were obtained using \proglang{Python}~3.10.2 with the
\pkg{openTSNE}~v0.6.1 package. \opentsne\ has three direct dependencies:
\pkg{numpy}, \pkg{scipy}, and \pkg{scikit-learn}. In our experiments, we used
\pkg{numpy}~v1.21.3, \pkg{scipy}~v1.7.1. \pkg{scikit-learn}~v1.0.1. As these
packages include their own downstream dependencies, the full environment used to
produce the figures is included in the accompanying GitHub repository
\url{https://github.com/pavlin-policar/opentsne-paper}.

>>>>

% General problem statement
One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data points~\citep{becht2019dimensionality}.
The culprit for slow response time stems from particular optimization procedures
and their specific implementation in popular
open-source libraries. Until quite recently, most popular implementations of
t-SNE were based on either a direct implementation of the t-SNE algorithm with
asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation
scheme developed by \citet{van2014accelerating} with asymptotic time complexity
$\mathcal{O}(N \log N)$. Recently, however, \citet{linderman2019fast} developed
a novel approximation scheme -- FIt-SNE -- which reduces the asymptotic time
complexity to $\mathcal{O}(N)$, overcoming the difficulties of scaling t-SNE
to large data sets.

% We benchmark Python
Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python t-SNE
implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE}
(v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne\ (v0.6.2). Benchmarks were
performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.
Benchmarks were run for $1,000$ iterations with the original t-SNE parameters,
as some implementations do not allow for their modification. To ensure a
realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse
brain data set~\citep{xxx}, containing 1.3 million single-cell gene-expression
profiles. We preprocess the data using the standard single-cell analysis
pipeline~\citep{XXX} and extract the top 50 principal components. To generate
benchmark data sets of different sample sizes, we subsample the data set five
times at each sample size. In total, we run each implementation on 30 different
data sets.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{benchmarks_python-final}
  \caption{\label{fig:benchmarks_py}
  We benchmark \opentsne\ (v0.6.2) against three popular open-source
  implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
  (v1.0.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
  on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
  that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
  same interpolation-based approximation scheme, while \pkg{scikit-learn} and
  \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
}
\end{figure*}

% Problem with Python
In Python, the most widely-used implementation of t-SNE comes from \pkg{scikit-learn},
which exhibits long runtimes when compared to other Python implementations.
\pkg{scikit-learn}, like its C++ counterpart -- \pkg{MulticoreTSNE}~\citep{Ulyanov2016} ,
is based on the Barnes-Hut approximation scheme. However, as
shown on Fig.~\ref{fig:benchmarks_py}, \pkg{scikit-learn} exhibits much longer
runtimes than \pkg{MulticoreTSNE}.
Newer implementations, such as \pkg{FIt-SNE} and \opentsne, both implement the
FIt-SNE approximation scheme, making them suitable for the visualization of
millions of data points. While both libraries implement the same
algorithm, \opentsne\ emphasizes ease of use and extensibility, and is primarily
written in Python, which inevitably introduces some runtime overhead. This
makes \opentsne\ marginally slower than its pure C++ counterpart -- \pkg{FIt-SNE}.
However, these differences disappear almost entirely when
utilizing multi-threading, as shown on Fig.~\ref{fig:benchmarks_py}.
Additionally, \opentsne\ provides a flexible API, allowing the user to split up
the embedding-construction process into several parts. This allows us to cache
and reuse intermediate results, enabling users to quickly experiment with
different parameter settings and iterate on their final visualizations. This
kind of functionality is not supported in other implementations, which require
the user to start from scratch for every embedding, resulting in large amounts
of redundant computation.

% Problem with other programming languages
In other programming languages, the selection of fast, open-source
implementations of t-SNE is more limited. In R, the most widely used
implementation of t-SNE -- \pkg{Rtsne}~\citep{krijthe2015rtsne} -- is based on
the Barnes-Hut approximation, and is closely comparable to 
\pkg{MulticoreTSNE}. In Julia, the most popular implementation -- 
\pkg{TSne.jl}~\citep{julia_tsne} -- is based on a naive $\mathcal{O}(N^2)$ implementation of the
algorithm, which allows it to be used only on smaller data sets containing up
to thousands of data points. We benchmark \pkg{Rtsne} (v0.15) and 
\pkg{TSne.jl} (v1.3.0) and compare it to \opentsne\ (v0.6.2).
Fig.~\ref{fig:benchmarks_lang} illustrates the scaling of each approximation 
scheme. Of these implementations, only \opentsne\ is able to scale to data sets
containing millions of data points in a reasonable amount of time.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{benchmarks_langs-final}
  %\includegraphics[width=\textwidth]{benchmarks-combined}
  \caption{\label{fig:benchmarks_lang}
  We select the fastest available, open-source implementations of the t-SNE
  algorithm from the Python, R, and Julia programming languages. We choose
  \opentsne\ (v0.6.2) for Python, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15)
  for R, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for Julia. The benchmarks
  reflect the time complexity of the
  implemented approximation schemes. For instance, \pkg{TSne.jl} implements a
  naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all
  but the smallest of data sets. \pkg{Rtsne} scales similarily to
  \pkg{MulticoreTSNE} in Python, as it implements the $\mathcal{O}(N \log N)$
  Barnes-Hut approximation scheme. Interestingly, the multi-threaded
  implementation version of \pkg{Rtsne} crashes on benchmark data sets
  containing over 500,000 data points due to lack of stack memory. We found
  these crashes to be inconsistent and somwewhat random. \opentsne\ is the only
  library implementing the more recent FIt-SNE approximation scheme, which makes
  it suitable for larger data sets.
}
\end{figure*}

@@

\begin{tabular}{l c c c c|c c}
\toprule
\setlength\tabcolsep{6pt}
& \rot{\pkg{scikit-learn}} & \rot{\pkg{MulticoreTSNE}} & \rot{\pkg{FIt-SNE}} & \rot{\pkg{openTSNE}} & \rot{\pkg{Rtsne}} & \rot{\pkg{Tsne.jl}} \\

\toprule
\textbf{Ease of Installation} \\
Native Python installation & \checkmark & & & \checkmark & & \\
\pkg{PyPI} package & \checkmark & \checkmark & & \checkmark & & \\
\pkg{conda} package & \checkmark & & & \checkmark & & \\

\hline
\textbf{Approximation Schemes} \\
None ($\mathcal{O}(N^2)$) & & & & & & \checkmark \\
Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark & \checkmark & \\
FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark & & \\

\hline
\textbf{Advanced Features and Extensions} \\
%Multiscale Gaussian kernels & & & & \checkmark & \checkmark & & \\
Extensible affinity models & & & & \checkmark & & \\
Variable degrees of freedom & & & \checkmark & \checkmark & & \\
Variable exaggeration & & & \checkmark & \checkmark & & \\
Globally-consistent initialization & & & \checkmark & \checkmark & & \\
Automatic learning rate & & & \checkmark & \checkmark & & \\
Adding samples to embeddings & & & & \checkmark & & \\
Callback system & & & & \checkmark & & \\
Interactive optimization  & & & & \checkmark & & \\

\hline
\textbf{Project Quality}\\
Actively maintained & \checkmark & & & \checkmark & \checkmark & \checkmark \\
Continuous integration & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
User documentation & \checkmark & & & \checkmark & \checkmark  & \\
End-to-end usage examples & \checkmark & & \checkmark & \checkmark & \checkmark & \checkmark \\
API reference & \checkmark & & & \checkmark & \checkmark &  \\
\bottomrule
\end{tabular}
\end{center}

\caption{\label{tab:features}
  We compare the features of \opentsne~(v0.6.2) to \pkg{scikit-learn}~(v1.0.1),
  \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
  \pkg{FIt-SNE}~(v1.1.0), as well as the major implementations in R and Julia,
  \pkg{Rtsne}~(v0.15), \pkg{Tsne.jl}~(v1.3.0). Active maintenance indicates whether the project has had any meaningful development in the past year from the time of writing.
}
\end{table}

@@

\section*{Reproducibility}

The results in this paper were obtained using \proglang{Python}~3.10.2 with the
\pkg{openTSNE}~v0.6.2 package. All results and figures can be reproduced using
the scripts found in the accompanying GitHub repository
\url{https://github.com/pavlin-policar/opentsne-paper}.
